The Calculus of Linear Constructions

### 2021-10-18 writing/lics22/lics22.tex
lics22

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{7pt}
\setlength{\grammarindent}{7em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{$\text{let }#1\text{ := }#2\text{ in }#3$}
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\utype}{:_{\scriptscriptstyle U}}
\newcommand{\ltype}{:_{\scriptscriptstyle L}}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\erase}[1]{\llbracket #1 \rrbracket}

@@

      <$m, n, A, B$> ::= $U_i$ | $L_k$ | $x$ \hspace*{4em} expressions
      \indalt $(x :_s A) \rightarrow B$
      \indalt $(x :_s A) \multimap B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \label{syntax}
  \end{figure}
  The syntax of the core type theory is presented in Figure \ref{syntax}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $k$ to specifically quantify over levels $0, 1, 2, ...$ that correspond to the predicative universes. We use the meta variable $i$ to quantify over all levels $*, 0, 1, 2, ...$. Here, $U_*$ is the impredicative universe of propositions, in the same spirit as CC$\omega$'s $Prop$ universe. $U_k$, and $L_k$ are the predicative universes of non-linear and linear types respectively.
  \begin{figure}[H]
    \vspace{-1.4em}
    \caption{Correspondence of CLC types and MELL implications}
    \Description{}
    \vspace{-1em}
    \begin{align}
      (\_ \utype A) \rightarrow B \quad &\Leftrightarrow \quad !(!A \multimap B) \\
      (\_ \ltype A) \rightarrow B \quad &\Leftrightarrow \quad !(A \multimap B) \\
      (\_ \utype A) \multimap B \quad &\Leftrightarrow \quad !A \multimap B \\
      (\_ \ltype A) \multimap B \quad &\Leftrightarrow \quad A \multimap B
    \end{align}
    \vspace{-1.6em}
    \label{correspondence}
  \end{figure}

@@

  \section{Extensions}
  \subsection{Inductive Types}
  Readers familiar with the literature on linear logic or linear type theory will have noticed at this point that CLC does not possess the famous $\otimes$ connective. Though it is entirely possible to encode $\otimes$ in core CLC through a linear Church encoding, proofs and programs written this way will be incredibly unwieldy.

  A major design goal of CLC is to support an ergonomic interface for users to define inductive types \cite{inductive,cic}. The presence of universe sorts $U$ and $L$ allows users to define inductive types to be non-linear or linear. The lack of ! exponential simplifies the construction and elimination of inductive terms through direct use of constructors and pattern-matching. The $\otimes$ connective becomes a definable inductive type.

  \begin{figure}[h]
  \vspace{-1.4em}
  \caption{Inductive Type Definition and Elimination}
  \begin{lstlisting}
  (* Sigma Definition *)
  Inductive Sigma (A: Type) (F: A -> Type): Type :=
  | pair : (x: A) -> F x -> Sigma A F.

@@

  Due to the complexity of checking soundness for dependent linear type definitions (such as \texttt{FTensor}) and pattern-matching based elimination, we have not verified the algorithm employed by our implementation yet.

  % \section{CLC in Action}

  % \subsection{Inductive Types}
  % A core design goal of CLC is to support an ergonomic interface for users to define both non-linear and linear inductive types. Inductive data should be both easy to construct, and to eliminate. The ! exponential, or its adjoint alternatives often get in the way of directly using data. Due to CLC's usage of universe sorts instead of !, data construction and elimination becomes much.

  % \subsection{Resource Reasoning}
  % \subsection{Parametric Reasoning}
  % \subsection{Mutable Data Structures}
  % \subsection{Continuations}
  % \subsection{Program Synthesis}

>>>>

\renewcommand{\syntleft}{}
\renewcommand{\syntright}{}
\setlength{\grammarparsep}{7pt}
\setlength{\grammarindent}{7em}
\newcommand{\indalt}[1][2]{\\\hspace*{-1.2em}\textbar\quad}
\newcommand{\rname}[1]{\textsc{\footnotesize #1}}
\newcommand{\pure}[1]{|#1|}
\newcommand{\sort}{\textbf{sort}}
\newcommand{\type}{\textbf{type}}
\newcommand{\refl}{\text{refl}}
\newcommand{\letin}[3]{\ $\text{let }#1\text{ := }#2\text{ in }#3$\ }
\newcommand{\ind}{\ $\text{ind}$\ }
\newcommand{\iter}{\ $\text{iter}$\ }
\newcommand{\new}{\text{new}}
\newcommand{\free}{\text{free}}
\newcommand{\get}{\text{get}}
\newcommand{\set}{\text{set}}
\newcommand{\utype}{:_{\scriptscriptstyle U}}
\newcommand{\ltype}{:_{\scriptscriptstyle L}}
\newcommand{\stype}[1]{:_#1}
\newcommand{\step}{\leadsto}
\newcommand{\pstep}{\leadsto}
\newcommand{\mrg}[3]{#1\ddagger#2\ddagger#3}
\newcommand{\erase}[1]{\llbracket #1 \rrbracket}

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{4em} expressions
      \indalt $(x :_s A) \rightarrow B$
      \indalt $(x :_s A) \multimap B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \label{syntax}
  \end{figure}
  The syntax of the core type theory is presented in Figure \ref{syntax}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $k$ to specifically quantify over levels $0, 1, 2, ...$ that correspond to the predicative universes. We use the meta variable $i$ to quantify over all levels $*, 0, 1, 2, ...$. Here, $U_*$ is the impredicative universe of propositions, in the same spirit as CC$\omega$'s $Prop$ universe. $U_k$, and $L_k$ are the predicative universes of non-linear and linear types respectively.
  \begin{figure}[H]
    \vspace{-1.4em}
    \caption{Correspondence of CLC types and MELL implications}
    \Description{}
    \vspace{-1em}
    \begin{align}
      (\_ \utype A) \rightarrow B \quad &\Leftrightarrow \quad !(!A \multimap B) \\
      (\_ \ltype A) \rightarrow B \quad &\Leftrightarrow \quad !(A \multimap B) \\
      (\_ \utype A) \multimap B \quad &\Leftrightarrow \quad !A \multimap B \\
      (\_ \ltype A) \multimap B \quad &\Leftrightarrow \quad A \multimap B
    \end{align}
    \vspace{-1.6em}
    \label{correspondence}
  \end{figure}

@@

  \section{Extensions}
  Readers familiar with the literature on linear logic or linear type theory will have noticed at this point that CLC does not possess the famous $\otimes$ connective. Though it is entirely possible to encode $\otimes$ in core CLC through a linear Church encoding, proofs and programs written in this style will be incredibly difficult to use.

  A major design goal of CLC is to support an ergonomic interface for users to define inductive types \cite{inductive,cic} as an extension to the core theory. The presence of universe sorts $U$ and $L$ allows users to declare inductive types to be non-linear or linear. The lack of ! exponential simplifies the construction and elimination of inductive terms through direct use of constructors and pattern-matching. The $\otimes$ connective becomes definable as an inductive type.

  \begin{figure}[h]
  \vspace{-1.2em}
  \caption{Inductive Type Definition and Elimination}
  \begin{lstlisting}
  (* Sigma Definition *)
  Inductive Sigma (A: Type) (F: A -> Type): Type :=
  | pair : (x: A) -> F x -> Sigma A F.

@@

  Due to the complexity of checking soundness for dependent linear type definitions (such as \texttt{FTensor}) and pattern-matching based elimination, we have not verified the algorithm employed by our implementation.

  Figure \ref{formation}, Figure \ref{introduction} and Figure \ref{elimination} presents the formation, introduction and elimination rules for select inductive types respectively. We will refer to these inductive types during the construction of examples.

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Formation Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash \mathbf{1} : U_0 }
      \rname{$1$-Form}

      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash \mathbb{N} : U_0 }
      \rname{$\mathbb{N}$-Form}

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A : U_i \\
        \Gamma \vdash m : A \\
        \Gamma \vdash n : A }
      { \Gamma \vdash m =_A n : U_* }
      \rname{=-Form}

      \inferrule
      { \pure{\Gamma} \\ 
        \Gamma \vdash A : U_k \\ 
        \Gamma x \utype A \vdash B : U_k  }
      { \Gamma \vdash \Sigma x:A.B : U_k }
      \rname{$\Sigma$-Form}

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A : L_k \\
        \Gamma \vdash B : L_k }
      { \Gamma \vdash A \otimes B : L_k }
      \rname{$\otimes$-Form}

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A : U_k \\ 
        \Gamma x \utype A \vdash B : L_k }
      { \Gamma \vdash F x:A.B : L_k }
      \rname{$F$-Form}
    \end{mathpar}
    \vspace{-0.4em}
    \label{formation}
    \Description{}
  \end{figure}

  Of all the rules presented here, perhaps the most surprising to readers familiar with linear types or dependent types is the \rname{$\mathbb{N}$-L-Elim} rule. This rule constructs a term with linear type through primitive recursion. The $\multimap$ arrows prevent duplication of the O-case element of linear type $C\ O$, whereas the S-case function is of non-linear type $(n : \mathbb{N}) \rightarrow C\ n \rightarrow C\ (S\ n)$ because it must be applied repeatedly during iteration.

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Introduction Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash () : \mathbf{1} }
      \rname{$1$-Intro}

      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash O : \mathbb{N} }
      \rname{O-Intro}

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash n : \mathbb{N} }
      { \Gamma \vdash S\ n : \mathbb{N} }
      \rname{S-Intro}

      \inferrule
      { \pure{\Gamma} \\ 
        \Gamma \vdash A : U_i \\
        \Gamma \vdash m : A }
      { \Gamma \vdash \refl_A m : m =_A m }
      \rname{=-Intro}

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash m : A \\
        \Gamma \vdash n : B[m/x] \\\\
        \Gamma \vdash \Sigma x:A.B : U_k }
      { \Gamma \vdash (m, n) : \Sigma x:A.B }
      \rname{$\Sigma$-Intro}

      \inferrule
      { \Gamma_1 \vdash m : A \\
        \Gamma_2 \vdash n : B \\ 
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\\\
        \Gamma \vdash A \otimes B : L_k }
      { \Gamma \vdash (m, n) : A \otimes B }
      \rname{$\otimes$-Intro}

      \inferrule
      { \pure{\Gamma_1} \\
        \Gamma_1 \vdash m : A \\
        \Gamma_2 \vdash n : B[m/x] \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
        \Gamma \vdash F x:A.B : L_k }
      { \Gamma \vdash (m, n) : F x:A.B }
      \rname{$F$-Intro}
    \end{mathpar}
    \vspace{-2em}
    \label{introduction}
    \Description{}
  \end{figure}

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Elimination Rules}
    \begin{mathpar}
      \inferrule
      { \Gamma_1 \vdash m : \mathbf{1} \\ 
        \Gamma_2 \vdash n : A \\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{()}{m}{n} : A }
      \rname{$1$-Elim}

      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash \iter_\mathbb{N}^U :
        (C \utype \mathbb{N} \rightarrow U_i) \rightarrow C\ O \rightarrow \\
        ((n \utype \mathbb{N}) \rightarrow C\ n \rightarrow C\ (S\ n)) \rightarrow
        (n \utype \mathbb{N}) \rightarrow C\ n }
      \rname{$\mathbb{N}$-U-Elim}

      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash \iter_\mathbb{N}^L :
        (C \utype \mathbb{N} \rightarrow L_k) \rightarrow C\ O \multimap \\
        ((n \utype \mathbb{N}) \rightarrow C\ n \rightarrow C\ (S\ n)) \multimap
        (n \utype \mathbb{N}) \multimap C\ n }
      \rname{$\mathbb{N}$-L-Elim}

      \inferrule
      { \Gamma_1 \vdash m : \Sigma x:A.B \\ 
        \Gamma_2, x \utype A, y \utype B \vdash n : C \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{$(x, y)$}{$m$}{$n$} \vdash C }
      \rname{$\Sigma$-Elim}

      \inferrule
      { \Gamma_1 \vdash m : A \otimes B \\ 
        \Gamma_2, x \ltype A, y \ltype B \vdash n : C \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{$(x, y)$}{$m$}{$n$} \vdash C }
      \rname{$\otimes$-Elim}

      \inferrule
      { \Gamma_1 \vdash m : F x:A.B \\
        \Gamma_2, x \utype A, y \ltype B \vdash n : C \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} }
      { \Gamma \vdash \letin{$(x, y)$}{$m$}{$n$} \vdash C }
      \rname{$F$-Elim}
    \end{mathpar}
    \vspace{-0.8em}
    \label{elimination}
    \Description{}
  \end{figure}

  \section{Applications}
  In this section, we demonstrate some of the possible applications of CLC through examples. We collapse the sort annotations on arrows $(x \utype A) \rightarrow B$ and $(x \ltype A) \rightarrow B$ into $(x : A) \rightarrow B$ whenever $A$'s sort is inferable from context.

  \subsection{Stateful Reasoning}
  We first postulate five state axioms. These axioms could be viewed as an interface to a trusted memory allocator.
  \begin{definition}
    State axioms.
    \begin{itemize}
      \item $l \mapsto_A m : \mathbb{N} \rightarrow (A : U_k) \rightarrow A \rightarrow L_k$ \\
      The $\mapsto$ is a linear type constructor. It fulfills a role similar to L3's \cite{l3} memory access capability, or Separation Logic's \cite{reynolds02} points-to assertion. Intuitively, it is a proof that a term $m$ of non-linear type $A$ is currently stored at address $l$.
      \item $\new : (A : U_k) \rightarrow (m : A) \rightarrow F l:\mathbb{N}.(l \mapsto_A m)$ \\
      $\new$ is a function that allocates a new memory store for term $m$ of non-linear type $A$. It results in an address $l$ paired with an assertion $l \mapsto_A m$.
      \item $\free : (A : U_k) \rightarrow F l:\mathbb{N}.F m:A.(l \mapsto_A m) \rightarrow \mathbf{1}$ \\
      $\free$ is a function that de-allocates the memory store at location $l$ which is currently in use. This can be seen in its input $F l:\mathbb{N}.F m:A.(l \mapsto_A m)$, where the $(l \mapsto_A m)$ component is a proof that location $l$ is currently allocated.
      \item $\get : (A : U_k) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \rightarrow F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$ \\
      $\get$ is a function that retrieves the term stored at address $l$ when given a proof $(l \mapsto_A m)$ that there is indeed a term $m$ currently stored there. The type of $\get$'s output is the rather complicated $F \_: (\Sigma x. x =_A m). (l \mapsto_A m)$. This not only returns the term stored at $l$, but also a proof that the returned term is exactly the same as what $(l \mapsto_A m)$ claimed. The proof $(l \mapsto_A m)$ is returned unchanged as well.
      \item $\set : (A\ B : U_k) \rightarrow (l : \mathbb{N}) \rightarrow (m : A) \rightarrow (l \mapsto_A m) \multimap (n : B) \multimap (l \mapsto_B n)$ \\
      $\set$ has the most complicated type of all the axioms presented here. Intuitively, if address $l$ is allocated and storing some term $m$ of type $A$, $\set$ may put term $n$ of type $B$ into the the store at address $l$, overwriting the original stored term. This is the so called "strong update". The consumption of $(l \mapsto_A m)$ and return of $(l \mapsto_B n)$ encode this update process.
    \end{itemize}
  \end{definition}
